TacticToe: Learning to Prove with Tactics

### 2018-03-21 tactictoe_jv.tex
Yutaka's comments

\section{Introduction}
Many of the state-of-the-art interactive theorem provers (ITPs) such as
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight},
  \isabelle~\cite{isabelle}
  and \coq~\cite{coq-book} provide high-level parameterizable tactics for constructing proofs.
  Tactics analyze the current proof state (goal and
  assumptions) and apply non-trivial proof transformations.
  Formalized proofs take advantage of different levels of automation:
  specialized rules, theory-based strategies and general purpose strategies.
  Thanks to progress in proof
  automation, developers can delegate more and more complicated proof
  obligations to general purpose strategies. Those are implemented by automated
  theorem provers (ATPs) such as \eprover~\cite{eprover}. Communication
  between
  an ITP and ATPs is made possible by a ``hammer''
  system~\cite{hammers4qed,tgck-cpp15}. It acts as an interface by performing
  premise selection, translation and proof reconstruction.
  Yet, ATPs are far from foolproof and more precise user-guidance, achieved
  by applying a particular sequence of specialized rules, is almost always
  necessary to develop a mathematical theory.

  In this work, we develop in the ITP \holfour a procedure to select suitable
  tactics depending on the current proof state by learning
  from previous proofs. Backed by this machine-learned guidance, our prover
  \tactictoe relies on \emph{Monte
  Carlo tree search}~\cite{montecarlo} to find sequences of tactic applications
  leading to an ITP proof.

\subsection{The problem}
An ITP is a development environment for the construction of formal proofs in 
which it is possible, but not common, to write a proof as a sequence of 
applications of primitive inference rules.
The common approach, however, is to use and devise high-level tools and 
automation that abstract over useful ways of producing proof pieces.
A specific, prevalent instance of this approach is the use of \emph{tactics} 
for \emph{backward} or goal-directed proof.
Here, the ITP user operates on a proof state, initially the desired conjecture, 
and applies tactics that transform the proof state until it is solved (or until 
giving up).
Each tactic performs a sound transformation of the proof state: essentially, it 
promises that if the resulting proof state can be solved, then so can the 
starting one.
This gives rise to a machine learning problem: can we learn a mapping from a 
given proof state to the next tactic (or sequence of tactics) that will 
productively advance the proof towards a solution?

\paragraph{Goals and theorems.}
A goal (or proof state) is represented as a \emph{sequent}.
A sequent is composed of a set of assumptions and a conclusion, all of which 
are higher-order logic formulas~\cite{gordon01hol}.
When a goal is proven, it becomes a theorem.
If the developer gives the theorem a name then we refer to it as a 
\textit{top-level} theorem.

\paragraph{Theories and script files.}
Formal developments in \holfour are organized into named theories, which are
concretely implemented by \emph{script} files defining modules in the
functional programming language \sml.
Each script file corresponds to a single theory, and contains definitions of
types and constants as well as statements of theorems together with their
proofs.
In practice, tactic-based proofs are written in an informally specified
\emph{embedded domain-specific language}: the language consisting of
pre-defined tactics and \emph{tacticals} (functions that operate on tactics).
However, the full power of \sml is always available for users to define their
own tactics or combinations thereof.
%To automate proof search strategies a user can implement programming language functions that correspond to more involved proofs.
%Functions that correspond to backward proofs, called tactics, are combined with each other with the help of tacticals (functions that takes
%tactics as arguments).
%Such a combination of tactics will be called a \emph{proof script} when it proves a theorem stated in a theory.

\paragraph{Tactics in \holfour.}
Considering concrete implementation details now, a \emph{tactic} is a function
that takes a goal (or proof obligation) and returns a list of goals (subgoals
of the original goal) together
with a \textit{validation function}.
Calling the validation function on a list of \emph{theorems}, 
corresponding to
the list of subgoals, results in a theorem corresponding to the original goal.
For example, if the list of subgoals is empty, then calling the 
validation
function on the empty list yields the originally goal sequent as a theorem.
In this way, tactics implement the plumbing of backwards, goal-directed, proof
construction.
Because validation functions are only important to check the final 
proof, we
omit them during the description of the proof search algorithm. We denote by
$t(g)$ the list of goals produced by the application of a tactic $t$ to a goal
$g$.

\subsection{Contributions}
This paper extends the work described in \cite{tgckju-lpar17}.
There we proposed the idea of a tactic-based prover based on supervised
learning guidance. And we achieved a 39\% success rate on theorems of the
\holfour standard library by running \tactictoe with a 5 seconds timeout.
The contributions of this extension and their effect on our system are:

\begin{itemize}
\item Proof recording at the tactic level is made more precise in 
Section~\ref{sec:recording}. Pattern
matching constructions, opening of submodules are supported. The different
recording steps (parsing, unfolding and replaying) are regrouped into a single
\sml function.
\item Monte Carlo tree search (MCTS) replaces A* as our
search algorithm in Section~\ref{sec:proofsearch}. The MCTS algorithm gives a 
more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by predictors presented in 
Section~\ref{s:prediction} for
three kinds of objects: tactics, theorems and lists of goals.
\item The orthogonalization process that eliminates redundant tactics is
described in a precise manner in Section~\ref{sec:ortho}.
\item A tactic abstraction mechanism is introduced in 
Section~\ref{sec:synthesis}. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover in Section~\ref{sec:atp} to help \tactictoe 
during proof search.
\item Evaluation of \tactictoe with a 60 seconds timeout achieving a 66\%
success rate on the standard library. And detailed comparisons between 
\tactictoe and \eprover on the different type of problems are reported in 
Section~\ref{s:experiments}.
\item Minimization and embellishment of the discovered proof facilitates user
interaction in Section~\ref{sec:proofdisplay}.
\end{itemize}

@@

In this section, we only give a description of the graph at a given
moment of the search after a certain amount of tactic applications.
Construction of the tree is done by the MCTS algorithm in
Section~\ref{sec:proofsearch}, which is guided by prediction algorithms
presented in Section~\ref{s:prediction}.

\begin{definition}\label{def:stree}(search tree)\\
A search tree $\mathfrak{T}$ is a sixtuple
$(\mathbb{T},\mathbb{G},\mathbb{A},T,G,A)$
that respects the following conditions:
\begin{itemize}
\item $\mathbb{T}$ is a set of tactics, $\mathbb{G}$ is a set of goals
 and $\mathbb{A}$ is a set of nodes representing lists of goals. All objects 
 are tagged with their position in the tree (see Figure~\ref{fig:choice}).
\item $T$ is a function from $\mathbb{G}$ to $\mathcal{P}(\mathbb{T})$. It
takes a goal and return the set of tactics already applied to this goal.
\item $G$ is a function from $\mathbb{A}$ to $\mathcal{P}(\mathbb{G})$.
It takes a node and returns the list of goals of this node.
\item $A$ is a function that takes a pair $(g,t)\in \mathbb{G} \times 
\mathbb{T}$ such as $t \in T(g)$ and 
returns a node $A(g,t)$ such as $t(g) = G(A(g,t))$. In other words, the output 
$t(g)$ is exactly the list of goals contained in $A(g,t)$.
\item It is acyclic, i.e. a node cannot be a strict descendant (see 
Definition~\ref{def:desc}) of itself.
\item There is exactly one node with no parents. This root node contains 
exactly one goal which is the initial goal (also called conjecture).
\end{itemize}

@@

If no explicit order is given, we assume that the sets
$\mathbb{T},\mathbb{G},\mathbb{A}$ are equipped with an
arbitrary \textit{total order}. In Figure~\ref{fig:choice}, a part of the a 
search tree is depicted, with goals represented by circles, nodes by rectangles 
and tactics by arrows.

@@

We call a goal (respectively node) \textit{unsolved} if it is not in the set of 
solved
goals (respectively nodes).
The \textit{open goal} of a node is the first unsolved goal of this node 
according to
some preset order. The other unsolved nodes are called pending nodes.
If a node is unsolved, then it contains a unique open goal.
During MCTS exploration, we do not explore pending goals of an unsolved node
before its open goal is solved which justifies our terminology. In other words,
our search tree has the following property: if $g$ is a pending node then
$T(g)$ is empty.

@@

\begin{definition}\label{def:prod} (productive tactic)
The application of a tactic $t$ on an open goal $g$ is called 
\textit{productive} if and only if
all the following conditions are satisfied:
\begin{itemize}
\item It does not fail or timeout. To prevent tactics from looping, we
interrupt tactics after a short amount of time (0.05 seconds).
\item It does not loop, i.e., its output does not contain any goals that appear
in the ancestor of the node of the current goal.
\item It is not a parallel step, i.e., there does not exist $t'$ in $T(g)$ 
such as $t'(g) \subseteq t(g)$.
\end{itemize}

@@

\section{Prediction}\label{s:prediction}
The learning-based selection of relevant lemmas significantly improves the
automation for hammers~\cite{BlanchetteGKKU16}. In \tactictoe we use the
distance-weighted \emph{$k$ nearest-neighbour} classifier~\cite{DudaniS76}
adapted for lemma selection~\cite{ckju-pxtp13}. It allows for hundreds of
predictions per second, while maintaining very high prediction quality~\cite{femalecop}.
We rely on this supervised machine learning algorithm to predict three
different kinds of objects:
 tactics, theorems and lists of goals. The number of predicted objects $k$ is
 called the \emph{prediction radius}. For a goal $g$, we denote
by $\mathit{Predictions}^{\mathit{tactic}}_k (g)$ (respectively
$\mathit{Predictions}^{\mathit{theorem}}_k (g)$
and $\mathit{Predictions}^{\mathit{goal\_list}}_k (g)$) the $k$ tactics
(respectively theorems and lists of goals) selected by our prediction algorithm.

@@

\paragraph{Tactics}
To start with, we build a database of tactics consisting
of tactic-goal pairs recorded from successful tactic applications in human
proofs (recording will be discussed in
Section~\ref{sec:recording}).
During proof search, the recorded goals will be reordered according to their
similarity with a target goal $g$ (usually the first pending goal of a node).
The recorded pairs and their similarity to $g$ induce an
order on tactics. The intuition is that tactics which have been successful on
goals similar to $g$ should be tried first, since they are more likely to lead
to a proof.
This predicted tactic order is translated into a prior policy for our MCTS
algorithm (see Section~\ref{sec:policy}).
Tactic selection is also used to improve the quality of the database
of tactics during orthogonalization (see Section~\ref{sec:ortho}).

@@

\paragraph{Lists of Goals as Output of Tactics}
A dataset of tactic outputs is compiled during orthogonalization (see 
Section~\ref{sec:ortho}).
Some elements of this set will be considered positive examples (see 
Section~\ref{sec:evaluation}).
During proof search, given a list of
goals $l$ created by a tactic, we select a set of lists of goals that are most
similar to $l$. The ratio of positive examples in the selection gives us a
prior evaluation for MCTS.

@@

We predict tactics through their associated goals. So, the features
for each kind of object can be extracted from (different representations of)
mathematical formulas. We start by describing features for \holfour terms. From
there, we extend the extraction mechanism
to goals, theorems and lists of goals. Duplicated features are always removed
so that each object has an associated set of features.

Here are the different kinds of feature we extract from terms:
\begin{itemize}
\item names of constants, including the logical operators,
\item type constructors present in the types of constants and variables,
\item subterms with all variables replaced by a single placeholder $V$,
\item names of the variables.
\end{itemize}
Goals (respectively theorems) are represented by pairs consisting of a list of
terms (assumptions) and a term (conclusion). We distinguish between
features of the assumptions and features of the conclusion by adding a
different tag to elements of each set. The feature of a goal (respectively 
theorem)
are the union of all these tagged features. From
that, we can construct features for a list of goals by computing the union of 
the features of each goal in the list.

We estimate the similarity (or co-distance) between two objects $o_1$ and $o_2$
through their respective feature sets $f_1$ and $f_2$.
The estimation is based on the frequency of the features in the intersection of
$f_1$ and $f_2$. A good rule of thumb is that the rarer the shared features
are, the more similar two goals should be. The relative rarity of each feature
can be estimated by calculating its TF-IDF weight~\cite{Jones72astatistical}.
We additionally raise these weights to the sixth power giving
even more importance
to rare features. This modification was found experimentally optimal in
\cite{ckju-pxtp13}. \todoi {Cezary: Josef says sixth power is not in this 
citation}.

The first co-distance $sim_1$ sums the weight of all shared
features to compute the total score.
In a second co-distance $sim_2$, we additionally take
the total number of features into account, to reduce the seemingly unfair
advantage of big
feature sets in the first scoring function.
\[sim_1 (o_1, o_2) = {\sum\nolimits_{\,f \in f_0 \cap 
f_1}{\text{tfidf}(f)^{6}}}\]
\[sim_2 (o_1, o_2) = \frac{{\sum\nolimits_{\,f \in f_0 \cap
f_1}{\text{tfidf}(f)^{6}}}}
{ln (e + \mathit{card} f_0 + \mathit{card} f_1)}\]

@@

Despite the best efforts of the prediction algorithms, a selected tactic may
not solve the current goal, proceed in the wrong direction
or even loop. For that reason, predictions need to be accompanied by a
proof search mechanism that allows for backtracking and
can choose which proof tree to extend next and in which direction.
Our search algorithm is a Monte Carlo tree search~\cite{montecarlo} algorithm
that relies
on a prior evaluation function and a prior policy function. Those priors are
learned through direct supervised learning for the policy and via the data
accumulated during orthogonalization for the evaluation. Therefore, this MCTS
algorithm does not need to rely on roll outs for evaluation. The
policy and evaluation are estimated by the simple supervised k-NN algorithm
here instead of neural networks trained through reinforcement learning
as e.g. in AlphaGo Zero~\cite{silver2017mastering}.
A characteristic of the MCTS algorithm is that it offers a good trade-off
between exploitation and exploration. This means that the algorithm searches
deeper the more promising branches and leaves enough time for the exploration
of less likely alternatives.

@@

The proof search starts by creating an initial search tree. The search tree
then evolves by the repetitive applications of MCTS steps. A step in the main
loop of MCTS consists of three parts: node selection, node extension, and
backpropagation. And the decision to terminate the algorithm is taken during
resolution.

@@

%A similar trade-off between exploration and exploitation exists in algorithm
%like Monte Carlo. However, Monte Carlo tree search metrics are based on many
%simulations and do not reflect our exploration strategy that only
\paragraph{Node selection}
Through node extension steps the search tree grows and the number of paths to
explore increases. To decide which node to extend next, an evolving value for
each node is given in Definition~\ref{def:value}. The algorithm that performs
node selection starts from the root of the search tree. From the current node,
the function \texttt{CompareChildren} chooses among the children of its open
goal the one with the highest value.

If the highest children value is higher than the \textit{widening policy} (see 
Section~\ref{sec:policy}), then the
selected child
becomes the current node, otherwise the final result of node selection is the
current node. The node selection algorithm can be written as the following
pseudo-code:

@@

\begin{definition}\label{def:value}(Value)
The value of the $i^{th}$ child $a_i$ of the open goal of a parent node $p$
is determined by:
\[\mathit{Value}(a_i) = \mathit{CurEvaluation}(a_i) + c_{exploration} *
\mathit{Exploration}(a_i)\]

The number $\mathit{Failure}(a_i)$ is the number of failures that occurred
during node extension from descendants of $a_i$ .
The current evaluation $CurEvaluation(a_i)$ is the average evaluation of
all descendants of $a_i$ including node extension failures.

@@

@@

\[\mathit{CurPolicy}(a_i) = \frac{1 +
\mathit{Visit}(a_i)}{\sqrt{\mathit{Visit}(p)}}\]
The policy is approximately the  percentage of time a node was visited. The
square root skews this probability to favor even more exploration of nodes with
few visits.

The coefficient $c_{exploration}$ is experimentally determined and adjusts the
trade-off between exploitation and exploration.
\end{definition}

\paragraph{Node extension}
Let $a$ be the result of node selection.
If $a$ is a solved node or the descendant of a solved node, then extending $a$
is useless and the algorithm reports a failure for this MCTS step.
%In our
%experiments, filtering such nodes before node selection is detrimental.
If $a$ is not $solved$, it applies the best untested tactic $t$ on the open
goal $g$ of this
node according to the prediction order for $g$. If no such tactic exists or if
$t$ is not productive, a failure is returned.
If node extension succeeds, a new node containing $t(g)$ is added to the
children of $a$.

@@

\paragraph{Resolution}
The search ends when the algorithm reaches one of these 3 conditions:
\begin{itemize}
\item It finds a proof, i.e., the root node is solved. In this case,
the search returns a minimized and
prettified proof script (see Section~\ref{sec:proofdisplay}).
\item It saturates, i.e., there are no tactics to be applied to any open goal.
This occurs less than 10 times in the full-scale experiment.
This happens only at the beginning of the \holfour library, where the training 
data
has very few tactics.
\item It exceeds a given time limit (default is set to 60 seconds).
\end{itemize}

\subsection{Supervised learning guidance}
Our MCTS algorithm relies on the guidance of two supervised learning methods.
The prior policy decides the order in which tactic should be tried and the
prior evaluation estimates if a branch is fruitful by analysing tactic
outputs. Both priors influence the rates at which branches of the search tree
are explored.

>>>>

\section{Introduction}
Many of the state-of-the-art interactive theorem provers (ITPs) such as
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight},
  \isabelle~\cite{isabelle}
  and \coq~\cite{coq-book} provide high-level parameterizable tactics for constructing proofs.
  Tactics analyze the current proof state (goal and
  assumptions) and apply non-trivial proof transformations.
  Formalized proofs take advantage of different levels of automation which are 
  in increasing order of generality:
  specialized rules, theory-based strategies and general purpose strategies.
  Thanks to progress in proof
  automation, developers can delegate more and more complicated proof
  obligations to general purpose strategies. Those are implemented by automated
  theorem provers (ATPs) such as \eprover~\cite{eprover}. Communication
  between
  an ITP and ATPs is made possible by a ``hammer''
  system~\cite{hammers4qed,tgck-cpp15}. It acts as an interface by performing
  premise selection, translation and proof reconstruction.
  Yet, ATPs are not flawless and more precise user-guidance, achieved
  by applying a particular sequence of specialized rules, is almost always
  necessary to develop a mathematical theory.

  In this work, we develop in the ITP \holfour a procedure to select suitable
  tactics depending on the current proof state by learning
  from previous proofs. Backed by this machine-learned guidance, our prover
  \tactictoe executes a \emph{Monte
  Carlo tree search}~\cite{montecarlo} algorithm to find sequences of tactic 
  applications
  leading to an ITP proof.

\subsection{The problem}
An ITP is a development environment for the construction of formal proofs in 
which it is possible to write a proof as a sequence of 
applications of primitive inference rules. This approach is not preferred 
by developers because proving a valuable theorem requires in general more than 
thousand primitive inferences.
The common approach, however, is to use and devise high-level tools and 
automation that abstract over useful ways of producing proof pieces.
A specific, prevalent instance of this approach is the use of \emph{tactics} 
for \emph{backward} or goal-directed proof.
Here, the ITP user operates on a proof state, initially the desired conjecture, 
and applies tactics that transform the proof state until it is solved.
Each tactic performs a sound transformation of the proof state: essentially, it 
promises that if the resulting proof state can be solved, then so can the 
starting one. This gives rise to a machine learning problem: can we learn a 
mapping from a given proof state to the next tactic (or sequence of tactics) 
that will productively advance the proof towards a solution?

\paragraph{Goals and theorems.}
A goal (or proof state) is represented as a \emph{sequent}.
A sequent is composed of a set of assumptions and a conclusion, all of which 
are higher-order logic formulas~\cite{gordon01hol}.
When a goal is proven, it becomes a theorem.
And when the developer has given the theorem a name, we refer to it as a 
\textit{top-level} theorem.

\paragraph{Theories and script files.}
Formal developments in \holfour are organized into named theories, which are
implemented by \emph{script} files defining modules in the
functional programming language \sml.
Each script file corresponds to a single theory, and contains definitions of
types and constants as well as statements of theorems together with their
proofs.
In practice, tactic-based proofs are written in an informally specified
\emph{embedded domain-specific language}: the language consisting of
pre-defined tactics and \emph{tacticals} (functions that operate on tactics).
However, the full power of \sml is always available for users to define their
own tactics or combinations thereof.
%To automate proof search strategies a user can implement programming language functions that correspond to more involved proofs.
%Functions that correspond to backward proofs, called tactics, are combined with each other with the help of tacticals (functions that takes
%tactics as arguments).
%Such a combination of tactics will be called a \emph{proof script} when it proves a theorem stated in a theory.

\paragraph{Tactics in \holfour.}
A \emph{tactic} is a function
that takes a goal (or proof obligation) and returns a list of goals (subgoals
of the original goal) together
with a \textit{validation function}.
Calling the validation function on a list of \emph{theorems}, 
corresponding to
the list of subgoals, results in a theorem corresponding to the original goal.
For example, if the list of subgoals is empty, then calling the 
validation
function on the empty list yields the originally goal sequent as a theorem.
In this way, tactics implement the plumbing of goal-directed proof
construction, transforming the conjecture by reasoning in a backward manner.
Since validation functions are only important to check the final 
proof, we
omit them during the description of the proof search algorithm. We denote by
$t(g)$ the list of goals produced by the application of a tactic $t$ to a goal
$g$.

\subsection{Contributions}
This paper extends the work described in our previous work~\cite{tgckju-lpar17}.
There we proposed the idea of a tactic-based prover based on supervised
learning guidance. And we achieved a 39\% success rate on theorems of the
\holfour standard library by running \tactictoe with a 5 seconds timeout.
The contributions of this paper and their effect on our system are:

\begin{itemize}
\item Proof recording at the tactic level is made more precise in 
Section~\ref{sec:recording}. Pattern
matching constructions, opening of submodules are supported. 
%The different
%recording steps (parsing, unfolding and replaying) are regrouped into a single
%\sml function.
\item Monte Carlo tree search (MCTS) replaces A* as our
search algorithm in Section~\ref{sec:proofsearch}. The MCTS algorithm gives a 
more balanced feedback on
the success of
each proof step by
comparing subtrees of the search tree instead of leafs. The policy and
evaluation are learned
through supervised learning.
\item Proof guidance required by MCTS is given by prediction algorithms 
through supervised learning (see 
Section~\ref{s:prediction}).
These predictors are implemented for three kind of objects: tactics, 
theorems and lists of goals.
\item The orthogonalization process that eliminates redundant tactics is
described in a precise manner in Section~\ref{sec:ortho}.
\item A tactic abstraction mechanism is introduced in 
Section~\ref{sec:synthesis}. It enables us to create
more general and flexible tactics where tactic arguments are dynamically
predicted.
\item The internal ATP \metis is complemented
by asynchronous calls to \eprover in Section~\ref{sec:atp} to help \tactictoe 
during proof search.
\item Evaluation of \tactictoe with a 60 seconds timeout achieves a 66\%
success rate on the standard library. And comparisons between 
\tactictoe and \eprover on different type of problems are reported in 
Section~\ref{s:experiments}.
\item Minimization and embellishment of the discovered proof facilitates user
interaction in Section~\ref{sec:proofdisplay}.
\end{itemize}

@@

In this section, we only give a description of the graph at a given
moment of the search after a certain amount of tactic applications.
Construction of the tree is done by the MCTS algorithm in
Section~\ref{sec:proofsearch}. The MCTS algorithm is guided by prediction 
algorithms presented in Section~\ref{s:prediction}.

\begin{definition}\label{def:stree}(search tree)\\
A search tree $\mathfrak{T}$ is a sextuple
$(\mathbb{T},\mathbb{G},\mathbb{A},T,G,A)$
that respects the following conditions:
\begin{itemize}
\item $\mathbb{T}$ is a set of tactics, $\mathbb{G}$ is a set of goals
 and $\mathbb{A}$ is a set of nodes representing lists of goals. All objects 
 are tagged with their position in the tree (see Figure~\ref{fig:choice}).
\item $T$ is a function from $\mathbb{G}$ to $\mathcal{P}(\mathbb{T})$. It
takes a goal and return the set of tactics already applied to this goal.
\item $G$ is a function from $\mathbb{A}$ to $\mathcal{P}(\mathbb{G})$.
It takes a node and returns the list of goals of this node.
\item $A$ is a function that takes a pair $(g,t)\in \mathbb{G} \times 
\mathbb{T}$ such as $t \in T(g)$ and 
returns a node $A(g,t)$ such as $t(g) = G(A(g,t))$. In other words, the output 
$t(g)$ is exactly the list of goals contained in $A(g,t)$.
\item It is acyclic, i.e. a node cannot be a strict descendant (see 
Definition~\ref{def:desc}) of itself.
\item There is exactly one node with no parents. This root node contains 
exactly one goal which is the initial goal (also called conjecture).
\end{itemize}

@@

If no explicit order is given, we assume that the sets
$\mathbb{T},\mathbb{G},\mathbb{A}$ are equipped with an
arbitrary \textit{total order}. In Figure~\ref{fig:choice}, a part of the a 
search tree is depicted. Goals, nodes and tactics are represented 
respectively by circles, rectangles and arrows.

@@

We call a goal \textit{unsolved} if it is not in the set 
of solved goals. Similarly, we call a node \textit{unsolved} if it does not 
belong to the set of solved nodes.
The \textit{open goal} of a node is the first unsolved goal of this node 
according to
some preset order. The other unsolved nodes are called \textit{pending nodes}.
If a node is unsolved, then it contains a unique open goal.
During MCTS exploration, we do not explore pending goals of an unsolved node
before its open goal is solved which justifies our terminology. In other words,
our search tree has the following property: if $g$ is a pending node then
$T(g)$ is empty.

@@

\begin{definition}\label{def:prod} (productive tactic)
The application of a tactic $t$ on an open goal $g$ is called 
\textit{productive} if and only if
all the following conditions are satisfied:
\begin{itemize}
\item It does not fail or timeout. To prevent tactics from looping, we
interrupt tactics after a short amount of time (0.05 seconds).
\item It does not loop, i.e., its output does not contain any goals that appear
in the ancestor of the node of the current goal.
\item It is not a redundant step, i.e., there does not exist $t'$ in $T(g)$ 
such as $t'(g) \subseteq t(g)$.
\end{itemize}

@@

\section{Prediction}\label{s:prediction}
The learning-based selection of relevant lemmas significantly improves the
automation in hammer systems~\cite{BlanchetteGKKU16}. In \tactictoe we use the
distance-weighted \emph{$k$ nearest-neighbour} classifier~\cite{DudaniS76}
adapted for lemma selection~\cite{ckju-pxtp13}. It allows for hundreds of
predictions per second, while maintaining very high prediction quality~\cite{femalecop}.
We rely on this supervised machine learning algorithm to predict objects of 
three kinds: tactics, theorems and lists of goals. The number of predicted 
objects $k$ is
 called the \emph{prediction radius}. For a goal $g$, we denote
by $\mathit{Predictions}^{\mathit{tactic}}_k (g)$ (respectively
$\mathit{Predictions}^{\mathit{theorem}}_k (g)$
and $\mathit{Predictions}^{\mathit{goal\_list}}_k (g)$) the $k$ tactics
(respectively theorems and lists of goals) selected by our prediction algorithm.

@@

\paragraph{Tactics}
To start with, we build a database of tactics consisting
of tactic-goal pairs recorded from successful tactic applications in human
proofs (recording will be discussed in
Section~\ref{sec:recording}).
During proof search, the recorded goals will be reordered according to their
similarity with a target goal $g$ (usually the first pending goal of a node).
The recorded pairs and their similarity to $g$ induce an
order on tactics. Intuitively, tactics which have been successful on
goals similar to $g$ should be tried first. Indeed, they are more likely to lead
to a proof.
This predicted tactic order is then transformed into a prior policy of our MCTS
algorithm (see Section~\ref{sec:policy}).
Tactic selection is also used to improve the quality of the database
of tactics during orthogonalization (see Section~\ref{sec:ortho}).

@@

\paragraph{Lists of Goals as Output of Tactics}
We compile a dataset of tactic outputs during orthogonalization (see 
Section~\ref{sec:ortho}).
Some elements of this set will be considered positive examples (see 
Section~\ref{sec:evaluation}).
During proof search, given a list of
goals $l$ created by a tactic, we select a set of lists of goals that are most
similar to $l$. The ratio of positive examples in the selection gives us a
prior evaluation for MCTS.

@@

We predict tactics through their associated goals. So, the features
for each kind of object can be extracted from different representations of
mathematical formulas. We start by describing features for \holfour terms. From
there, we extend the extraction mechanism
to goals, theorems and lists of goals. Duplicated features are always removed
so that each object has an associated set of features.

Here are the kinds of features we extract from terms:
\begin{itemize}
\item names of constants, including the logical operators,
\item type constructors present in the types of constants and variables,
\item subterms with all variables replaced by a single placeholder $V$,
\item names of the variables.
\end{itemize}
Goals (respectively theorems) are represented by pairs consisting of a list of
terms (assumptions) and a term (conclusion). We distinguish between
features of the assumptions and features of the conclusion by adding a
different tag to elements of each set. The feature of a goal (respectively 
theorem)
are the union of all these tagged features. From
that, we can construct features for a list of goals by computing the union of 
the features of each goal in the list.

We estimate the similarity between two objects $o_1$ and $o_2$
through their respective feature sets $f_1$ and $f_2$.
The estimation is based on the frequency of the features in the intersection of
$f_1$ and $f_2$. A good rule of thumb is that the rarer the shared features
are, the more similar two goals should be. The relative rarity of each feature
can be estimated by calculating its TF-IDF weight~\cite{Jones72astatistical}.
We additionally raise these weights to the sixth power giving
even more importance
to rare features. The second and third authors found this modification 
experimentally 
optimal~\cite{ckju-pxtp13}. \todoi {Cezary: Josef says sixth 
power is not 
in this 
citation}.

The first similarity measure $sim_1$ sums the weight of all shared
features to compute the total score.
In a second similarity measure $sim_2$, we additionally take
the total number of features into account, to reduce the seemingly unfair
advantage of big
feature sets in the first scoring function.
\[sim_1 (o_1, o_2) = {\sum\nolimits_{\,f \in f_0 \cap 
f_1}{\text{tfidf}(f)^{6}}}\]
\[sim_2 (o_1, o_2) = \frac{{\sum\nolimits_{\,f \in f_0 \cap
f_1}{\text{tfidf}(f)^{6}}}}
{ln (e + \mathit{card} f_0 + \mathit{card} f_1)}\]

@@

Our prediction algorithms are not always accurate. 
Therefore, a selected tactic may
fail, proceed in the wrong direction or even loop. 
For that reason, predictions need to be accompanied by a
proof search mechanism that allows for backtracking and
can choose which proof tree to extend next and in which direction.
Our search algorithm is a Monte Carlo tree search~\cite{montecarlo} algorithm
that relies
on a prior evaluation function and a prior policy function. Those priors are
learned through direct supervised learning for the policy and via the data
accumulated during orthogonalization for the evaluation. Therefore, this MCTS
algorithm does not need to rely on roll outs for evaluation. The
policy and evaluation are estimated by the simple supervised k-NN algorithm.
A characteristic of the MCTS algorithm is that it offers a good trade-off
between exploitation and exploration. This means that the algorithm searches
deeper the more promising branches and leaves enough time for the exploration
of less likely alternatives.

\begin{remark}
Neural networks trained through reinforcement learning can be 
very effective for approximating the policy and evaluation as demonstrated in 
.e.g. AlphaGo Zero~\cite{silver2017mastering}. 
But training neural networks is 
computationally expensive and has not yet been 
proven effective in our context. That is why we chose a simpler machine 
learning model for our project.
\end{remark}

@@

The proof search starts by creating a search tree during initialization. 
The search tree then evolves by the repetitive applications of MCTS steps. A 
step in the main loop of MCTS consists of three parts: node selection, node 
extension, and backpropagation. And the decision to stop the loop is taken 
by the resolution module.

@@

%A similar trade-off between exploration and exploitation exists in algorithm
%like Monte Carlo. However, Monte Carlo tree search metrics are based on many
%simulations and do not reflect our exploration strategy that only
\paragraph{Node selection}
Through node extension steps the search tree grows and the number of paths to
explore increases. To decide which node to extend next, the MCTS algorithm
computes for each node a value (see Definition~\ref{def:value}) that changes
after each MCTS step. 
The algorithm that performs
node selection starts from the root of the search tree. From the current node,
the function \texttt{CompareChildren} chooses the child node with the highest 
value among the children of its open goal.

If the highest children value is higher than the \textit{widening policy} (see 
Section~\ref{sec:policy}), then the
selected child
becomes the current node, otherwise the final result of node selection is the
current node. The following pseudo-code illustrates our node selection 
algorithm:

@@

\begin{definition}\label{def:value}(Value)\\
The value of the $i^{th}$ child $a_i$ of the open goal of a parent node $p$
is determined by:
\[\mathit{Value}(a_i) = \mathit{CurEvaluation}(a_i) + c_{exploration} *
\mathit{Exploration}(a_i)\]

The current evaluation $CurEvaluation(a_i)$ is the average evaluation of
all descendants of $a_i$ including node extension failures.

@@

where the number $\mathit{Failure}(a_i)$ is the number of failures that occurred
during node extension from descendants of $a_i$ .

@@

\[\mathit{CurPolicy}(a_i) = \frac{1 +
\mathit{Visit}(a_i)}{\sqrt{\mathit{Visit}(p)}}\]
The policy can be seen as a skewed percentage of the number of times a node was 
visited. The square root favors exploration of nodes with
few visits. The coefficient $c_{exploration}$ is experimentally determined and 
adjusts the trade-off between exploitation and exploration.
\end{definition}

\paragraph{Node extension}
Let $a$ be the result of node selection.
If $a$ is a solved node or the descendant of a solved node, then extending $a$
would not be productive and the algorithm reports a failure for this MCTS step.
If $a$ is not solved, it applies the best untested tactic $t$ on the open
goal $g$ of this
node according to the prediction order for $g$. 
If no such tactic exists or if $t$ is not productive, the algorithm returns a 
failure.
If node extension succeeds, a new node containing $t(g)$ is added to the
children of $a$.

@@

\paragraph{Resolution}
The search ends when the algorithm reaches one of these 3 conditions:
\begin{itemize}
\item It finds a proof, i.e., the root node is solved. In this case,
the search returns a minimized and
prettified proof script (see Section~\ref{sec:proofdisplay}).
\item It saturates, i.e., there are no tactics to be applied to any open goal.
This occurs less than 0.1 percent of the time in the full-scale experiment.
And this happens only at the beginning of the \holfour library, where the 
training 
data
has very few tactics.
\item It exceeds a given time limit (default is set to 60 seconds).
\end{itemize}

\subsection{Supervised learning guidance}
Our MCTS algorithm relies on the guidance of two supervised learning methods.
The prior policy estimates the probability of success of a tactic $t$ on a 
goal $g$ and associate it with the potential children that would be created by 
the 
application of $t$ to $g$. The prior evaluation judges if a branch is fruitful 
by analyzing tactic
outputs. Both priors influence the rates at which branches of the search tree
are explored.

\paragraph{Prior policy}\label{sec:policy}
Similarity between goals produced by the tactic predictor are hard
to translate into a prior policy, so we rely solely on the prediction order
to estimate the probability with which each tactic should be tried.
Let $\mathfrak{T}$ be a search tree after a number of MCTS steps.
Let $p$ be a selected node in $\mathfrak{T}$ and $g$ its open goal. We order
the list of $n$ productive tactics already
applied to $g$ by their similarity score. We note the resulting list 
$t_0,\ldots,t_{n-1}$.
Let $c_{policy}$ be a constant heuristic optimized during our experiments.
We calculate the prior policy of a child $a_i$ produced by the tactic
$t_i$ by:
\[PriorPolicy(a_i) = (1 - c_{policy})^{i} * c_{policy}\]
